Urban and socio-economic correlates of property prices in Dublin's area

### 2020-08-19 manuscript/manuscript.Rmd
final submission

```{r setup, include = FALSE}
# options ----------------------------------------------------------------------
base::set.seed(123)
base::options(
  scipen = 999,
  knitr.kable.NA = " - "
  )
knitr::opts_chunk$set(
  cache.extra = knitr::rand_seed, 
  message = FALSE, 
  warning = FALSE, 
  error = FALSE, 
  echo = FALSE,
  cache = TRUE,
  fig.retina = 3
  )

@@

Beside their inherent characteristics such as size, design and materials [@bourassa2007spatial; @liu2013spatial], property valuation is dictated by multiple external factors explaining their spatial autocorrelation [@basu1998analysis]. Some of these factors are macroeconomic [@antonakakis2016dynamic], related to the evolution of population general wealth while others are related to the characteristics of the neighbourhood [@dubin1998predicting]. These characteristics include the urban and socio economic structure of the neighbourhood [@goodman1998housing]. However, it is difficult to evaluate the influence of such external factors on property valuation [@clapp2002predicting]. By using machine learning algorithms, prediction of property prices has become more and more accurate. Techniques used usually include artificial neural networks due to the volatility of the market [@limsombunchai2004house; @feng2015comparing; @selim2009determinants]. Indeed, artificial neural network techniques can quickly adjust to volatility changes, therefore incorporating sharp increases or decreases in their predictions [@aragones2007neural]. However, the model obtained from most artificial neural networks are hardly interpretable [@rudin2019stop].

In this paper, an alternative machine learning algorithm is used to identify urban and socio-economic correlates of property price evolution in the area of Dublin, Ireland in 2018. Due to the high number of potential features, an eXtreme Gradient Boosting (XGBoost) regression model is believed to obtain the best results with low parameter customisation and high speed [@chen2016xgboost]. Contrary to most artificial neural network techniques, XGBoost is an interpretation-focused method which automatically provides estimates of feature importance from a trained predictive model.

@@

Since 2010, under the Irish Property Services (Regulation) Act, all individuals acquiring a property in Ireland have to declare it to Property Services Regulatory Authority (PSRA). They must provide details such as the date of sale, the price and the address of all residential properties purchased in Ireland as declared to the Revenue Commissioners for stamp duty purposes [@rppr2020]. Data is filed electronically by persons doing the conveyancing of the property on behalf of the purchaser, and it must be noted that errors may occur when the data is being filed. 
Because of the evolution of urban planning and the evolution of socio-economic features measured by the 2016 census, only properties sold in 2018 were processed to avoid potential temporal incoherence. The total amount of property sold in 2018 in the county Dublin which appears in the Property Price Register is `r n_raw`. From these, all property corresponding to apartments were removed (`r scales::percent(1-(n_house/n_raw), accuracy = 0.1)` of the original dataset). Finally, unsuccessfully geocoded properties were also removed from the original dataset (`r scales::percent(1-(n_final/n_house), accuracy = 0.1)`) to obtain the `r n_final` of the final dataset used to identify the urban and socio-economic correlates of property prices.

@@

To estimate the geographical distribution of the price density, multiple methods such as multiple regression [@mccluskey2000application] or Bayesian smoothing [@clapp2002predicting] have been employed. While these methods are efficient in a non-restricted space, they have limitations when they are used in coastal areas. This problem is known as "finite area smoothing" and occurs when predictions from a model are approximated across geographical barriers, such as irregular coastal shapes, which would lead to poorly predicted values [@ramsay2002spline]. To deal with the influence of geographical boundaries on price estimation, a generalised additive model using soap film smoothing has been used [@wood2008soap].

@@

In addition to the distance to every urban feature, each property sold in 2018 was related to 48 socio-economic features of the small area including the property, as measured by the Irish census 2016. The results of Irish 2016 census consultation is accessible through both the Central Statistics Office [@cso2020] and All-Island Research Observatory [@airo2020]. The data obtained can be mapped over small area boundaries which are fractions of Irish Electoral Division map. The social features extracted are corresponding to population information, religion, carers and health. Economic features correspond to the characteristics of each small area; including the proportion of housing types, number of rooms, occupancy and tenure per small area. Each property is then associated to the value corresponding to its small area. For anonymisation purposes, the results from small areas having less than five respondents to the census were converted to a proportion of five respondents.

@@

Both tree and linear regression based boosting model are used for every of the continuous output learning task in order to compare their accuracy with urban and socio-economic features. As no specific weight or bias in the models has been hypothesized, the parameters of the XGBoost function have been set to the default value. Then, the most accurate model is used to identify the features that are contributing the most to the prediction of property prices. This importance is a measure of the improvement in accuracy brought by a feature based on its coefficient's absolute magnitude.

@@

```{r distrib-plot, fig.height=4, out.width="0.98\\columnwidth", fig.pos ="!h", fig.cap="(ref:distrib-plot)"}
data_dublin_2018_features %>%
  magrittr::use_series(price) %>%
  stats::density() %$%
  data.frame(x, y) %>%
  ggplot() +
  geom_segment(aes(
    x = x,
    xend = x,
    y = 0,
    yend = y,
    color = x
  )) +
  scale_x_continuous("Price", labels = dollar_format(suffix = "", prefix = "€")) +
  scale_y_continuous("Density") +
  scale_color_gradient(
    low = '#3a53a4', 
    high = '#ed1c24', 
    ) +
  theme_apa() +
  theme(
    text = element_text(family = "serif", size = 12),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "none"
  )
```

```{r distrib-gps, fig.height=5, out.width="0.98\\columnwidth", fig.pos ="!h", fig.cap="(ref:distrib-gps)"}
gam_pred %>%
  dplyr::filter(value > 200000) %>%
  ggplot(aes(x = lng, y = lat)) + 
  geom_raster(aes(fill = value), na.rm = TRUE) +
  geom_contour(aes(z = value), color = "black", na.rm = TRUE) +
  geom_sf(data = dublin_sea, fill = "light blue", inherit.aes = FALSE) +
  scale_fill_gradient(
    "Price Density", 
    low = '#3a53a4', 
    high = '#ed1c24', 
    trans = "log",
    breaks = c(250000, 1000000), 
    labels = scales::dollar_format(suffix = "", prefix = "€"))  +
  scale_x_continuous(
    "Longitude", 
    limits = c(-6.5, -6), 
    breaks = seq(-6.5, -6, 0.2)) +
  scale_y_continuous(
    "Latitude", 
    limits = c(53.2, 53.5), 
    breaks = seq(53.2, 53.5, 0.1)) +
  theme_classic() +
  theme(
    text = element_text(family = "serif", size = 12),
    legend.position = "none",
    panel.background = element_rect(fill = "grey50")
  )
```

@@

```{r osm-features-xgb, fig.cap="(ref:osm-features-xgb)", fig.height=7, fig.pos ="!h", out.width="0.98\\columnwidth"}
result_accuracy <- result %>%
  ggplot(aes(x = price, y = pred)) +
  geom_point(alpha = 0.1) +
  geom_smooth(
    method = "lm",
    fullrange = TRUE,
    se = TRUE,
    color = "black",
    size = 0.1,
    na.rm = TRUE
  ) +
  scale_x_continuous(
    "Property Price",
    labels = scales::dollar_format(suffix = "", prefix = "€")
  ) +
  scale_y_continuous(
    "Predicted Property Price",
    labels = scales::dollar_format(suffix = "", prefix = "€")
  ) +
  theme_apa() +
  theme(
    text = element_text(family = "serif", size = 12)
  ) +
  ggtitle("A")

result_error <- result %>%
  dplyr::mutate(diff_percent = (pred - price) / price) %>%
  ggplot(aes(x = price, y = diff_percent)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(
    color = "black",
    se = TRUE,
    fullrange = TRUE,
    size = 0.1,
    na.rm = TRUE
  ) +
  scale_x_continuous(
    "Property Price",
    labels = scales::dollar_format(suffix = "", prefix = "€")
  ) +
  scale_y_continuous(
    "Predicted Property Price Error",
    limits = c(-0.5, 0.5),
    labels = scales::percent
  ) +
  theme_apa() +
  theme(
    text = element_text(family = "serif", size = 12)
  ) +
  ggtitle("B")

@@

By analysing their importance (Table \@ref(tab:osm-features-table)), the most relevant geographical features to predict housing prices are the distance to an embassy (`r table_osm$Importance[1]`) and the distance to natural grasslands such as parks and gardens (`r table_osm$Importance[2]`).

@@

```{r census-features-xgb, fig.cap="(ref:census-features-xgb)", fig.height=7, fig.pos ="!h", out.width="0.98\\columnwidth"}
result_accuracy <- result %>%
  ggplot(aes(x = price, y = pred)) +
  geom_point(alpha = 0.1) +
  geom_smooth(
    method = "lm",
    fullrange = TRUE,
    se = TRUE,
    color = "black",
    size = 0.1,
    na.rm = TRUE
  ) +
  scale_x_continuous(
    "Property Price",
    labels = scales::dollar_format(suffix = "", prefix = "€")
  ) +
  scale_y_continuous(
    "Predicted Property Price",
    labels = scales::dollar_format(suffix = "", prefix = "€")
  ) +
  theme_apa() +
  theme(text = element_text(family = "serif", size = 12)) +
  ggtitle("A")

result_error <- result %>%
  dplyr::mutate(diff_percent = (pred - price) / price) %>%
  ggplot(aes(x = price, y = diff_percent)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(
    color = "black",
    se = TRUE,
    fullrange = TRUE,
    size = 0.1,
    na.rm = TRUE
  ) +
  scale_x_continuous(
    "Property Price",
    labels = scales::dollar_format(suffix = "", prefix = "€")
  ) +
  scale_y_continuous(
    "Predicted Property Price Error",
    limits = c(-0.5, 0.5),
    labels = scales::percent
  ) +
  theme_apa() +
  theme(text = element_text(family = "serif", size = 12)) +
  ggtitle("B")

@@

While Dublin is a very specific city due to its low population density spread on a large surface, its housing market is one of the most expensive in Europe. The main reason is the structure of the market, mainly made of residential properties and few apartment buildings. In this study, the prices of `r n_final` properties sold in 2018 in the area of Dublin were analysed from the Property Services Regulatory Authority dataset in order to identify the potential factors correlating with these prices. The enforcement of a housing price registry has been shown as being an essential tool to regulate the housing market [@kitchin2013making]. However, a major limitation of this study is the absence of characteristics for the property itself. Indeed, prices are mainly determined by factors such as the size of the property or the number of rooms. In addition, while some features appear to be more important than others in their contribution to property prices, these results are not causal. Thus, features' importance can be either the cause or the consequence of property prices in a neighborhood.

@@

The evolution of housing prices is a pressing issue in most European capitals and specially in Dublin. Given their significant increase, properties are less and less affordable for individuals. By preforming a feature analysis with urban and socio-economic features, it is possible to evaluate and predict the potential price of a property. Indeed features such as the presence of embassies or parks are criteria that influence significantly the price of properties. Similarly, the characteristics of inhabitants in the area such as religion, health and age is correlated to the evolution of housing prices. These results allow an understanding of why some areas have higher prices than others which is a relevant information not only for real estate agents in charge of property valuations but also to buyers in order to approximate the real value of a property.

@@

The authors would like to thank the developers of the following R packages used to process, analyse, display and report data: `r papaja::cite_r(file = "bib/r-references.bib")`.

>>>>

```{r setup, include = FALSE}
# options ----------------------------------------------------------------------
base::set.seed(123)
base::options(
  scipen = 999,
  knitr.kable.NA = " - "
  )

knitr::opts_chunk$set(
  cache.extra = knitr::rand_seed, 
  message = FALSE, 
  warning = FALSE, 
  error = FALSE, 
  echo = FALSE,
  fig.retina = 3
  )

@@

Beside their inherent characteristics such as size, design and materials [@bourassa2007spatial; @liu2013spatial], property valuation is dictated by multiple external factors explaining their spatial autocorrelation [@basu1998analysis]. Some of these factors are macroeconomic [@antonakakis2016dynamic], related to the evolution of population general wealth while others are related to the characteristics of the neighborhood [@dubin1998predicting]. These characteristics include the urban and socio economic structure of the neighborhood [@goodman1998housing]. However, it is difficult to evaluate the influence of such external factors on property valuation [@clapp2002predicting]. By using machine learning algorithms, prediction of property prices has become more and more accurate. Techniques used usually include artificial neural networks due to the volatility of the market [@limsombunchai2004house; @feng2015comparing; @selim2009determinants]. Indeed, artificial neural network techniques can quickly adjust to volatility changes, therefore incorporating sharp increases or decreases in their predictions [@aragones2007neural]. However, the model obtained from most artificial neural networks are difficult to interpret [@rudin2019stop].

In this paper, an alternative machine learning algorithm is used to identify urban and socio-economic correlates of property price evolution in the area of Dublin, Ireland in 2018. Due to the high number of potential features, an eXtreme Gradient Boosting (XGBoost) regression model is believed to obtain the best results with low parameter customization and high speed [@chen2016xgboost]. Contrary to most artificial neural network techniques, XGBoost is an interpretation-focused method which automatically provides estimates of feature importance from a trained predictive model.

@@

Since 2010, under the Irish Property Services (Regulation) Act, all individuals acquiring a property in Ireland have to declare it to Property Services Regulatory Authority (PSRA). They must provide details such as the date of sale, the price and the address of all residential properties purchased in Ireland as declared to the Revenue Commissioners for stamp duty purposes [@rppr2020]. Data is filed electronically by persons doing the conveyancing of the property on behalf of the purchaser, and it must be noted that errors may occur when the data is being filed. 
Because of the evolution of urban planning and the evolution of socio-economic features measured by the 2016 census, only properties sold in 2018 were processed to avoid potential temporal incoherence. In 2018, the Property Price Register references `r n_raw` properties sold in Dublin's area. Among these properties, those corresponding to apartments have been removed from the original dataset in order to evaluate only houses (`r scales::percent(1-(n_house/n_raw), accuracy = 0.1)` of the properties). Finally, some properties have been unsuccessfully geocoded because of mistakes on their address or because they have not been found (`r scales::percent(1-(n_final/n_house), accuracy = 0.1)` of the properties). These properties have also been removed from the original dataset to obtain a final dataset of `r n_final` properties.

@@

To estimate the geographical distribution of the price density, multiple methods such as multiple regression [@mccluskey2000application] or Bayesian smoothing [@clapp2002predicting] have been employed. While these methods are efficient in a non-restricted space, they have limitations when they are used in coastal areas. This problem is known as "finite area smoothing" and occurs when predictions from a model are approximated across geographical barriers, such as irregular coastal shapes, which can lead to poorly predicted values [@ramsay2002spline]. To deal with the influence of geographical boundaries on price estimation, a generalized additive model using soap film smoothing has been used [@wood2008soap].

@@

In addition to the distance to every urban feature, each property sold in 2018 was related to 48 socio-economic features of the small area including the property, as measured by the Irish census 2016. The results of Irish 2016 census consultation is accessible through both the Central Statistics Office [@cso2020] and the All-Island Research Observatory [@airo2020] websites. The data obtained can be mapped over small area boundaries which are fractions of Irish Electoral Division map. The social features extracted are corresponding to population information, religion, carers and health. Economic features correspond to the characteristics of each small area; including the proportion of housing types, number of rooms, occupancy and tenure per small area. Each property is then associated to the value corresponding to its small area. For anonymisation purposes, the results from small areas having less than five respondents to the census were converted to a proportion of five respondents.

@@

Both tree and linear regression based boosting model are used for every of the continuous output learning task in order to compare their accuracy with urban and socio-economic features. As no specific weight or bias in the models have been hypothesized, parameters of the XGBoost algorithms have been kept as default values. Next, the most accurate model is used to identify the features that are contributing most to the prediction of property prices. This contribution (also called "importance") is a measure of the improvement in accuracy brought by a feature. It is based on the absolute magnitude of feature's coefficient.

@@

```{r distrib-plot, fig.height=4, out.width="0.98\\columnwidth", fig.pos ="!h", fig.cap="(ref:distrib-plot)"}
data_dublin_2018_features %>%
  magrittr::use_series(price) %>%
  stats::density() %$%
  data.frame(x, y) %>%
  ggplot() +
  geom_segment(aes(
    x = x,
    xend = x,
    y = 0,
    yend = y,
    color = x
  )) +
  scale_x_continuous("Price", labels = dollar_format(suffix = "", prefix = "€")) +
  scale_y_continuous("Density") +
  scale_color_gradient(low = '#3a53a4', high = '#ed1c24') +
  theme_apa() +
  theme(
    text = element_text(family = "serif", size = 14),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "none"
  )
```

```{r distrib-gps, fig.height=5, out.width="0.98\\columnwidth", fig.pos ="!h", fig.cap="(ref:distrib-gps)"}
gam_pred %>%
  dplyr::filter(value > 200000) %>%
  ggplot(aes(x = lng, y = lat)) + 
  geom_raster(aes(fill = value), na.rm = TRUE) +
  geom_contour(aes(z = value), color = "black", na.rm = TRUE) +
  geom_sf(data = dublin_sea, fill = "light blue", inherit.aes = FALSE) +
  scale_fill_gradient(
    "Price Density", 
    low = '#3a53a4', 
    high = '#ed1c24', 
    trans = "log",
    breaks = c(250000, 1000000), 
    labels = scales::dollar_format(suffix = "", prefix = "€"))  +
  scale_x_continuous(
    "Longitude", 
    limits = c(-6.5, -6), 
    breaks = seq(-6.5, -6, 0.2)) +
  scale_y_continuous(
    "Latitude", 
    limits = c(53.2, 53.5), 
    breaks = seq(53.2, 53.5, 0.1)) +
  theme_classic() +
  theme(
    text = element_text(family = "serif", size = 14),
    legend.position = "none",
    panel.background = element_rect(fill = "grey50")
  )
```

@@

```{r osm-features-xgb, fig.cap="(ref:osm-features-xgb)", fig.height=9, fig.pos ="!h", out.width="0.98\\columnwidth"}
result_accuracy <- result %>%
  ggplot(aes(x = price, y = pred)) +
  geom_point(alpha = 0.1) +
  geom_smooth(
    method = "lm",
    fullrange = TRUE,
    se = TRUE,
    color = "black",
    size = 0.1,
    na.rm = TRUE
  ) +
  scale_x_continuous(
    "Property Price",
    labels = scales::dollar_format(suffix = "", prefix = "€")
  ) +
  scale_y_continuous(
    "Predicted Property Price",
    labels = scales::dollar_format(suffix = "", prefix = "€")
  ) +
  theme_apa() +
  theme(
    text = element_text(family = "serif", size = 14)
  ) +
  ggtitle("A")

result_error <- result %>%
  dplyr::mutate(diff_percent = (pred - price) / price) %>%
  ggplot(aes(x = price, y = diff_percent)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(
    color = "black",
    se = TRUE,
    fullrange = TRUE,
    size = 0.1,
    na.rm = TRUE
  ) +
  scale_x_continuous(
    "Property Price",
    labels = scales::dollar_format(suffix = "", prefix = "€")
  ) +
  scale_y_continuous(
    "Predicted Property Price Error",
    limits = c(-0.5, 0.5),
    labels = scales::percent
  ) +
  theme_apa() +
  theme(
    text = element_text(family = "serif", size = 14)
  ) +
  ggtitle("B")

@@

By analyzing their importance (Table \@ref(tab:osm-features-table)), the most relevant geographical features to predict housing prices are the distance to an embassy (`r table_osm$Importance[1]`) and the distance to natural grasslands such as parks and gardens (`r table_osm$Importance[2]`).

@@

```{r census-features-xgb, fig.cap="(ref:census-features-xgb)", fig.height=9, fig.pos ="!h", out.width="0.98\\columnwidth"}
result_accuracy <- result %>%
  ggplot(aes(x = price, y = pred)) +
  geom_point(alpha = 0.1) +
  geom_smooth(
    method = "lm",
    fullrange = TRUE,
    se = TRUE,
    color = "black",
    size = 0.1,
    na.rm = TRUE
  ) +
  scale_x_continuous(
    "Property Price",
    labels = scales::dollar_format(suffix = "", prefix = "€")
  ) +
  scale_y_continuous(
    "Predicted Property Price",
    labels = scales::dollar_format(suffix = "", prefix = "€")
  ) +
  theme_apa() +
  theme(text = element_text(family = "serif", size = 14)) +
  ggtitle("A")

result_error <- result %>%
  dplyr::mutate(diff_percent = (pred - price) / price) %>%
  ggplot(aes(x = price, y = diff_percent)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(
    color = "black",
    se = TRUE,
    fullrange = TRUE,
    size = 0.1,
    na.rm = TRUE
  ) +
  scale_x_continuous(
    "Property Price",
    labels = scales::dollar_format(suffix = "", prefix = "€")
  ) +
  scale_y_continuous(
    "Predicted Property Price Error",
    limits = c(-0.5, 0.5),
    labels = scales::percent
  ) +
  theme_apa() +
  theme(text = element_text(family = "serif", size = 14)) +
  ggtitle("B")

@@

While Dublin is a very specific city due to its low population density spread on a large surface, its housing market is one of the most expensive in Europe. The main reason is the structure of the market, mainly made of residential properties and few apartment buildings. In this study, the prices of `r n_final` properties sold in 2018 in the area of Dublin from the Property Services Regulatory Authority dataset were analyzed in order to identify the potential factors correlating with these prices. The enforcement of a housing price registry has been shown as being an essential tool to regulate the housing market [@kitchin2013making]. However, a major limitation of this study is the absence of characteristics for the property itself. Indeed, prices are mainly determined by factors such as the size of the property or the number of rooms. In addition, while some features appear to be more important than others in their contribution to property prices, these results are not causal. Thus, a feature's importance can be either the cause or the consequence of property prices in a neighborhood

@@

The evolution of housing prices is a pressing issue in most European capitals and specially in Dublin. Given their significant increase, properties are less and less affordable for individuals. By preforming a feature analysis with urban and socio-economic features, it is possible to evaluate and predict the potential price of a property. Indeed features such as the presence of embassies or parks are criteria that influence significantly the price of properties. Similarly, the characteristics of inhabitants in the area such as religion, health and age is correlated to the evolution of housing prices. These results allow an understanding of why some areas have higher prices than others which is relevant information not only for real estate agents in charge of property valuations but also for buyers in order to estimate the real value of a property.

@@

The authors would like to thank the developers of the following R packages used to process, analyze, display and report data: `r papaja::cite_r(file = "bib/r-references.bib")`.
